---
title: "`r params$report_title`"
author: "`r params$report_author`"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    self_contained: true
params:
  # Path to the freshly dropped CSV to inspect. This is the ONE knob a user sets.
  data_path: !r here::here("1_data", "master_micro_survey.csv")

  # --- Report Text (editable; leave "" to use the smart default) ---
  report_title: "Pre-flight Data Quality Report"
  report_author: "CERP Analytics"
  source_note: "Generated via CERP Analytics — pre-flight QA (0.00)"
---

<!--
================================================================================
 0.00 — PRE-FLIGHT DATA QUALITY REPORT + TEMPLATE RECOMMENDER
--------------------------------------------------------------------------------
 A "zero template" you render against ANY freshly dropped CSV *before* picking a
 visual. It profiles the columns, flags the usual field-data landmines, and
 recommends which production templates (3.01–3.18) the data can actually feed —
 including which column maps to which *_var param.

 This template is PER-DATASET, not per-config: it is deliberately NOT wired into
 the master knit (2.2). Render it on its own, from the project root in RStudio:

     rmarkdown::render(
       here::here("3_templates", "0.00_data_quality_report.Rmd"),
       params     = list(data_path = here::here("1_data", "master_micro_survey.csv")),
       output_dir = here::here("4_output")
     )

 Point `data_path` at any CSV in 1_data/. Field data (anything other than
 master_*.csv) stays local — 4_output/ and 3_templates/*.html are gitignored, so
 a babychecker QA report never reaches git, and nothing here calls any API.
================================================================================
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,          # QA report: show findings, not code
  warning = FALSE,
  message = FALSE,
  dev.args = list(bg = "#FAF9F6"),
  fig.width = 9,
  fig.height = 4,
  fig.retina = 3
)
```

```{r load-theme-and-profile}
library(tidyverse)

source(here::here("2_R", "2.0_setup_theme.R"), local = TRUE)  # palette + theme_cerp()
source(here::here("2_R", "2.5_helpers.R"), local = TRUE)       # cerp_profile(), etc.

# DELIBERATE EXCEPTION to the "always cerp_load()" convention: this report exists
# to surface the very problems cerp_load() silently scrubs (header/value
# whitespace), so it must see the RAW file. cerp_profile() only describes — it
# never mutates — so reading raw here is safe.
if (!file.exists(params$data_path)) {
  stop("Data file not found: ", params$data_path,
       "\nCheck the 'data_path' parameter.", call. = FALSE)
}
raw  <- readr::read_csv(params$data_path, show_col_types = FALSE, trim_ws = FALSE)
prof <- cerp_profile(raw)
cols <- prof$columns
ds   <- prof$dataset
```

```{r callout-style, results='asis'}
# House-style callout boxes. Colors are pulled from the live palette (cerp_cols,
# sourced above) so they track theme_colors.yml rather than re-hardcoding hexes.
cat(sprintf('
<style>
  .cerp-callout { padding: 0.7rem 1.05rem; margin: 0.55rem 0; border-radius: 6px;
                  border-left: 4px solid %1$s; background: %2$s; color: %3$s;
                  font-family: "Libre Franklin", sans-serif; font-size: 0.94rem; }
  .cerp-callout.warn { border-left-color: %4$s; background: #fbf3e3; }
  .cerp-callout.flag { border-left-color: %5$s; background: #fbeae4; }
  .cerp-callout.ok   { border-left-color: %6$s; background: #eaf3ef; }
  .cerp-callout .lbl { font-weight: 600; }
</style>',
  cerp_cols[["primary"]], cerp_cols[["box"]], cerp_cols[["text"]],
  cerp_cols[["sand"]], cerp_cols[["accent"]], cerp_cols[["sage"]]))
```

This is a **pre-flight check** of `` `r basename(params$data_path)` `` — run *before*
choosing a visualization. It describes each column, flags the common field-data
problems, and recommends which CERP templates the data can feed. It changes
nothing: fix anything below in the source data, then re-run.

## Dimensions & duplicates

```{r dimensions, results='asis'}
cat(sprintf(
  "The dataset has **%s rows** and **%s columns**.\n\n",
  format(ds$n_rows, big.mark = ","), ds$n_cols))

# Dataset-level flags as house-style callouts (only shown when they fire).
if (ds$single_row) {
  cat('<div class="cerp-callout flag"><span class="lbl">Single row.</span> ',
      'One data row is not enough for most charts — expect distributions, ',
      'trends, and group comparisons to be meaningless.</div>\n', sep = "")
}
if (ds$duplicate_rows > 0) {
  cat(sprintf('<div class="cerp-callout warn"><span class="lbl">%s duplicate row(s).</span> ',
              ds$duplicate_rows),
      'Fully identical rows can double-count in aggregates — confirm they are real ',
      'repeated observations, not an export/merge artifact.</div>\n', sep = "")
}
if (length(ds$empty_columns) > 0) {
  cat(sprintf('<div class="cerp-callout warn"><span class="lbl">Empty column(s):</span> %s. ',
              paste(ds$empty_columns, collapse = ", ")),
      'These carry no values and are ignored by the recommender.</div>\n', sep = "")
}
if (length(ds$header_whitespace) > 0) {
  cat(sprintf('<div class="cerp-callout warn"><span class="lbl">Header whitespace:</span> %s. ',
              paste(sprintf("'%s'", ds$header_whitespace), collapse = ", ")),
      'cerp_load() squishes these automatically at render time, but tidy the source ',
      'header to be safe.</div>\n', sep = "")
}
if (!ds$single_row && ds$duplicate_rows == 0 &&
    length(ds$empty_columns) == 0 && length(ds$header_whitespace) == 0) {
  cat('<div class="cerp-callout ok"><span class="lbl">No structural issues.</span> ',
      'No duplicate rows, empty columns, or header whitespace.</div>\n', sep = "")
}
```

## Column profile

```{r profile-table}
profile_display <- cols %>%
  transmute(
    Column        = column,
    Type          = type,
    `Missing %`   = missing_pct,
    `Distinct`    = n_distinct,
    `Whitespace`  = whitespace_issues,
    `Date-parse`  = ifelse(is.na(date_parse_rate), "—",
                           scales::percent(date_parse_rate, accuracy = 1)),
    `Outliers`    = ifelse(is.na(outlier_count), "—", as.character(outlier_count))
  )

knitr::kable(profile_display, align = "llrrrrr")
```

*Type is a deterministic best guess (numeric / categorical / date / id-like /
text / empty). "Whitespace" counts values with stray leading/trailing/internal
spaces; "Date-parse" is the share of non-missing values `cerp_parse_date()` can
read (shown for text/date columns); "Outliers" counts points beyond 1.5×IQR.*

## Missingness

```{r missingness-chart}
miss_df <- cols %>% filter(missing_pct > 0) %>% arrange(missing_pct)

if (nrow(miss_df) == 0) {
  cat("No missing values in any column.")
} else {
  # Inline house-style bar chart (reuses theme_cerp() + palette — the sensible
  # reuse here; there is no template viz_*() for a QA missingness bar).
  miss_df <- miss_df %>% mutate(column = factor(column, levels = column))
  ggplot(miss_df, aes(x = missing_pct, y = column)) +
    geom_col(fill = cerp_cols[["primary"]], width = 0.72) +
    geom_text(aes(label = paste0(missing_pct, "%")),
              hjust = -0.15, family = cerp_font_caption,
              size = 3.2, color = cerp_cols[["subtle"]]) +
    scale_x_continuous(limits = c(0, max(miss_df$missing_pct) * 1.12),
                       expand = expansion(mult = c(0, 0))) +
    labs(title = "Missing values by column",
         subtitle = "Percent of rows with no usable value",
         x = NULL, y = NULL,
         caption = params$source_note) +
    theme_cerp() +
    theme(panel.grid.major.y = element_blank())
}
```

## Per-column issue callouts

```{r column-callouts, results='asis'}
any_issue <- FALSE

for (i in seq_len(nrow(cols))) {
  r <- cols[i, ]
  notes <- character(0)

  if (isTRUE(r$type == "empty")) {
    notes <- c(notes, "column is entirely empty (no usable values)")
  }
  if (!is.na(r$missing_pct) && r$missing_pct >= 20) {
    notes <- c(notes, sprintf("%.1f%% missing — thin coverage for this field",
                              r$missing_pct))
  }
  if (isTRUE(r$whitespace_issues > 0)) {
    notes <- c(notes, sprintf("%d value(s) have stray whitespace (cerp_load scrubs this at render)",
                              r$whitespace_issues))
  }
  if (!is.na(r$date_parse_rate) && r$date_parse_rate > 0 && r$date_parse_rate < 0.9) {
    notes <- c(notes, sprintf("only %s of values parse as dates — mixed/ambiguous date formats",
                              scales::percent(r$date_parse_rate, accuracy = 1)))
  }
  if (!is.na(r$outlier_count) && r$outlier_count > 0) {
    notes <- c(notes, sprintf("%d value(s) beyond 1.5×IQR — inspect for entry errors or true extremes",
                              r$outlier_count))
  }

  if (length(notes) > 0) {
    any_issue <- TRUE
    cls <- if (isTRUE(r$type == "empty") ||
               (!is.na(r$missing_pct) && r$missing_pct >= 50)) "flag" else "warn"
    cat(sprintf('<div class="cerp-callout %s"><span class="lbl">%s</span> (%s): %s.</div>\n',
                cls, r$column, r$type, paste(notes, collapse = "; ")))
  }
}

if (!any_issue) {
  cat('<div class="cerp-callout ok"><span class="lbl">No column-level issues.</span> ',
      'No high missingness, whitespace, date-parse, or outlier flags.</div>\n', sep = "")
}
```

## Recommended visuals

```{r recommender}
# =============================================================================
# DETERMINISTIC TEMPLATE RECOMMENDER
# -----------------------------------------------------------------------------
# A fixed rule table mapping the column-type profile (+ light name hints) to the
# production templates that data can feed. Each rule that matches emits one row:
# the template id, why it fits, the suggested column -> *_var mapping, and a
# score used only for ranking (higher = more specific / stronger match). No
# randomness, no model — same input always yields the same recommendations.
# Suggestions only: the user confirms the column mapping in the template YAML.
# =============================================================================

# --- Role extraction from the profile ----------------------------------------
nd        <- setNames(cols$n_distinct, cols$column)
by_type   <- function(t) cols$column[cols$type == t]
num_cols  <- by_type("numeric")
cat_cols  <- by_type("categorical")
date_cols <- by_type("date")
id_cols   <- by_type("id-like")

nm_match  <- function(pool, rx) pool[grepl(rx, pool, ignore.case = TRUE)]
first     <- function(v) if (length(v)) v[[1]] else NA_character_

# Likert-pattern detector: cleaned levels overlap a rating lexicon, 3–7 levels.
likert_lex <- c("strongly disagree","disagree","somewhat disagree","neutral",
                "neither agree nor disagree","somewhat agree","agree","strongly agree",
                "strongly dissatisfied","dissatisfied","satisfied","very satisfied",
                "strongly satisfied","never","rarely","sometimes","often","always",
                "poor","fair","good","very good","excellent")
is_likert <- function(colname) {
  v <- unique(tolower(stringr::str_squish(as.character(raw[[colname]]))))
  v <- v[nzchar(v) & !is.na(v)]
  if (length(v) < 3 || length(v) > 7) return(FALSE)
  mean(v %in% likert_lex) >= 0.6
}

# Named-role sub-pools (deterministic; name hints only sharpen, never fabricate).
year_num   <- nm_match(num_cols, "year|yr|wave|round|period")
time_cols  <- union(date_cols, year_num)
target_num <- nm_match(num_cols, "target|goal|benchmark")
event_num  <- nm_match(num_cols, "event.?time|rel.?time|time.?to|relative")
stage_cats <- nm_match(cat_cols, "stage|step|phase|pipeline")
district   <- nm_match(c(cat_cols, id_cols, by_type("text")),
                       "district|region|province|tehsil|division|zone")
# A treatment flag is any 2-level column — categorical ("Treatment"/"Control")
# OR numeric (0/1). Name hints (treat/arm/…) sharpen the pick when present.
binary_cols <- cols$column[cols$n_distinct == 2 &
                           cols$type %in% c("categorical", "numeric")]
treat_cats <- {
  hinted <- nm_match(binary_cols, "treat|arm|group|condition|intervention|treated")
  if (length(hinted)) hinted else binary_cols
}
likert_cats <- cat_cols[vapply(cat_cols, is_likert, logical(1))]
entity_cats <- nm_match(c(cat_cols, id_cols),
                        "district|school|clinic|facility|entity|unit|region|name")
# "Plain" measures: numeric that isn't a target/event-time/year role.
measure_num <- setdiff(num_cols, c(target_num, event_num, year_num))
outcome_pref <- function() {
  hinted <- nm_match(measure_num, "score|outcome|rate|result|amount|index|value")
  first(if (length(hinted)) hinted else measure_num)
}

# --- Rule accumulator ---------------------------------------------------------
recs <- list()
add  <- function(id, name, why, mapping, score) {
  recs[[length(recs) + 1]] <<- data.frame(
    id = id, template = name, why = why,
    mapping = paste(sprintf("%s = %s", names(mapping), unlist(mapping)),
                    collapse = "; "),
    score = score, stringsAsFactors = FALSE)
}
has <- function(x) length(x) > 0 && !all(is.na(x))

# --- Rules (one per production template) --------------------------------------
# 3.01 dumbbell — 2-level group + before/after numerics
{
  before <- first(nm_match(measure_num, "base|pre|before|initial"))
  after  <- first(nm_match(measure_num, "end|post|after|final"))
  if (has(treat_cats) && !is.na(before) && !is.na(after) && before != after) {
    add("3.01", "Baseline vs endline (dumbbell)",
        "A treatment/control group with matched before/after measures.",
        list(group_var = first(treat_cats), before_var = before, after_var = after), 9)
  }
}
# 3.02 distribution shifts — group + one numeric
if (has(cat_cols) && has(measure_num)) {
  add("3.02", "Distribution shifts (density)",
      "A grouping column and a numeric outcome to compare distributions.",
      list(group_var = first(cat_cols), outcome_var = outcome_pref()), 5)
}
# 3.03 treatment effects (forest) — treatment 2-level + numeric outcome
if (has(treat_cats) && has(measure_num)) {
  add("3.03", "Treatment effects (forest plot)",
      "A treatment/control flag and a numeric outcome.",
      list(group_var = first(treat_cats), outcome_var = outcome_pref()), 7)
}
# 3.04 subgroup impacts — treatment + a second categorical + numeric
{
  sub <- first(setdiff(cat_cols, first(treat_cats)))
  if (has(treat_cats) && !is.na(sub) && has(measure_num)) {
    add("3.04", "Subgroup impacts (coefficient plot)",
        "A treatment flag, a subgroup dimension, and a numeric outcome.",
        list(group_var = first(treat_cats), subgroup_var = sub,
             outcome_var = outcome_pref()), 6)
  }
}
# 3.05 waffle — a single categorical (parts of a whole)
if (has(cat_cols)) {
  add("3.05", "Waffle chart (share of whole)",
      "A categorical status column shown as proportions.",
      list(status_var = first(cat_cols)), 3)
}
# 3.06 slopegraph — entity + time (>=2 points) + numeric
if (has(entity_cats) && has(time_cols) && has(measure_num)) {
  add("3.06", "Slopegraph (two-point change)",
      "An entity, a time column, and a value — change between two periods.",
      list(entity_var = first(entity_cats), time_var = first(time_cols),
           value_var = outcome_pref()), 5)
}
# 3.07 diverging bar — entity + Likert response
if (has(likert_cats)) {
  add("3.07", "Diverging bar (Likert sentiment)",
      "A Likert-scale response column, ideal for a diverging bar.",
      list(entity_var = if (has(entity_cats)) first(entity_cats) else first(cat_cols),
           response_var = first(likert_cats)), 8)
}
# 3.08 icon array — a single categorical proportion
if (has(cat_cols)) {
  add("3.08", "Icon array (frequency)",
      "A categorical column shown as an intuitive icon frequency.",
      list(category_var = first(cat_cols)), 3)
}
# 3.09 waterfall — stage-like categorical + numeric
if (has(stage_cats) && has(measure_num)) {
  add("3.09", "Waterfall (running total by stage)",
      "A stage/step column with one numeric amount per stage.",
      list(category_var = first(stage_cats), value_var = outcome_pref()), 7)
}
# 3.10 bump — entity + time (>=3 points) + numeric
{
  t <- first(time_cols)
  if (has(entity_cats) && !is.na(t) && nd[t] >= 3 && has(measure_num)) {
    add("3.10", "Bump chart (ranking over time)",
        "An entity tracked across 3+ time points on a numeric metric.",
        list(entity_var = first(entity_cats), time_var = t,
             value_var = outcome_pref()), 6)
  }
}
# 3.11 bullet — entity + value + target numeric
if (has(entity_cats) && has(measure_num) && has(target_num)) {
  add("3.11", "Bullet chart (value vs target)",
      "An entity with a value and an explicit target/benchmark column.",
      list(entity_var = first(entity_cats), value_var = outcome_pref(),
           target_var = first(target_num)), 7)
}
# 3.12 deviation bar — entity/category + numeric
if (has(entity_cats) && has(measure_num)) {
  add("3.12", "Deviation bar (vs reference)",
      "An entity/category with a numeric value to show deviation from a reference.",
      list(entity_var = first(entity_cats), value_var = outcome_pref()), 4)
}
# 3.13 ridgeline — multi-level categorical + numeric distribution
{
  grp <- first(cat_cols[nd[cat_cols] >= 3])
  if (!is.na(grp) && has(measure_num)) {
    add("3.13", "Ridgeline (distributions by group)",
        "A categorical with 3+ groups and a numeric distribution to compare.",
        list(group_var = grp, value_var = outcome_pref()), 5)
  }
}
# 3.14 calendar heatmap — a date column + numeric
if (has(date_cols) && has(measure_num)) {
  add("3.14", "Calendar heatmap (daily values)",
      "A real date column with a numeric value — daily intensity over a calendar.",
      list(date_var = first(date_cols), value_var = outcome_pref()), 7)
}
# 3.15 choropleth — district-like names + numeric
if (has(district) && has(measure_num)) {
  add("3.15", "Choropleth map (by district)",
      "A district/region name column and a numeric metric to shade.",
      list(region_var = first(district), value_var = outcome_pref()), 8)
}
# 3.16 event study — unit + time + event-time + treatment + numeric outcome (panel)
if (has(id_cols) && has(time_cols) && has(event_num) &&
    has(treat_cats) && has(measure_num)) {
  add("3.16", "Event study (dynamic effects)",
      "A panel with unit, calendar time, event-time, a treatment flag, and an outcome.",
      list(unit_var = first(id_cols), time_var = first(time_cols),
           event_time_var = first(event_num), treat_var = first(treat_cats),
           outcome_var = outcome_pref()), 9)
}
# 3.17 small multiples — facet categorical + x + y
{
  facet <- first(cat_cols[nd[cat_cols] >= 3])
  xv    <- first(time_cols)
  if (!is.na(facet) && !is.na(xv) && has(measure_num)) {
    add("3.17", "Small multiples (faceted trend)",
        "A faceting category, an x (usually time), and a numeric y.",
        list(facet_var = facet, x_var = xv, y_var = outcome_pref()), 5)
  }
}
# 3.18 heatmap matrix — two categorical/time axes + numeric
{
  two_cats <- union(cat_cols, time_cols)  # rows x cols; a time column makes a good axis
  if (length(unique(two_cats)) >= 2 && has(measure_num)) {
    add("3.18", "Heatmap matrix (category × category)",
        "Two categorical/time axes and a numeric cell value.",
        list(row_var = two_cats[[1]], col_var = two_cats[[2]],
             value_var = outcome_pref()), 4)
  }
}
```

```{r recommender-table, results='asis'}
if (length(recs) == 0) {
  cat('<div class="cerp-callout flag"><span class="lbl">No confident recommendation.</span> ',
      'The column profile did not match any template rule cleanly — check the ',
      'flags above (empty columns, all-text data, single row) and confirm the ',
      'right file was profiled.</div>\n', sep = "")
} else {
  rec_df <- do.call(rbind, recs)
  rec_df <- rec_df[order(-rec_df$score, rec_df$id), ]
  rec_out <- data.frame(
    Rank             = seq_len(nrow(rec_df)),
    Template         = sprintf("%s — %s", rec_df$id, rec_df$template),
    `Why it fits`    = rec_df$why,
    `Suggested mapping` = rec_df$mapping,
    check.names = FALSE, stringsAsFactors = FALSE)
  cat(knitr::kable(rec_out, format = "html", align = "llll", row.names = FALSE))
}
```

*Recommendations are deterministic suggestions from the column profile — always
confirm the column → `*_var` mapping in the chosen template's YAML header before
rendering. Absence of a template here means the required column roles were not
detected, not that the chart is impossible.*

<p style="font-family:'Libre Franklin',sans-serif; font-size:0.8rem; color:#8a9aa3; margin-top:2rem;">
`r params$source_note`
</p>
